### 2020-07-02 paper.tex
draft abstract (too long)

@@

\begin{abstract}
  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  both left and right-hand margins. Use 10~point type, with a vertical
  spacing of 11~points. The \textbf{Abstract} heading must be centered,
  bold, and in point size 12. Two line spaces precede the
  Abstract. The Abstract must be limited to one paragraph.
\end{abstract}

@@

This is the best paper~\citep{cusumano2019gen}.

@@

@@

@@

@@

\subsection{Formalism}
Let $(Z, \Sigma, \mu)$ be a measure space of pairs of traces of the model and proposal, where $\mu$ is $\sigma$-finite.
Consider a $\mu$-measurable function $h : Z \to Z$ that is an involution ($h(h(z)) = z$) and let $\mu \circ h^{-1}$ denote the pushforward of $\mu$ under $h$, such that $\mu \circ h^{-1}$ is $\sigma$-finite and such that $\mu \circ h^{-1}$ is absolutely continuous with respect to $\mu$.
% TODO can we simplify the conditions, using the fact that h is an involution?
Then, the Radon-Nikodym derivative exists, and is denoted $(d (\mu \circ h^{-1}) / d \mu)(z)$.

The goal is to prove stationarity with respect to the measure defined by density $p(z)$ with respect to $\mu$.
The move starts with state $z \in Z$, applies the involution $h$, and then accepts (returns $h(z)$ with probability $\alpha(z)$) and otherwise returns $z$.
Let $k_{z}$ denote the measure on new states induced by the involution $h$ and the previous state $z$:
\begin{equation}
k_{z}(A) = [h(z) \in A] \alpha(z) + [z \in A] (1 - \alpha(z))
\end{equation}
The stationarity condition is:
\begin{equation}
\int_Z p(z) k_{z}(A) \mu(dz) = p(A) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
Substituting in the definition of $k_z(A)$ and simplifying gives:
\begin{equation} \label{eq:stationarity-requirement}
\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z p(z) [z \in A] \alpha(z) \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
Note that for a $\mu$-measurable function $g$ such that $g \circ h$ is integrable with respect to $\mu$:
\begin{equation} \label{eq:pushforward}
\int_Z g(h(z)) \mu(dz) = \int_Z g(z) (\mu \circ h^{-1})(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
\end{equation}
For $g(z) := p(h(z)) [z \in A] \alpha(h(z))$ we have $g(h(z)) = p(z) [h(z) \in A] \alpha(z)$, which is the integrand of the left-hand-side of Equation~(\ref{eq:stationarity-requirement}).
Applying Equation~(\ref{eq:pushforward}) to the left-hand side of Equation~(\ref{eq:stationarity-requirement}) gives:
\begin{equation}
\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
\end{equation}
Equating this with the right-hand side of Equation~(\ref{eq:stationarity-requirement}) gives the following sufficient condition for stationarity:
\begin{equation}
\int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
It therefore suffices to find $\alpha$ such that:
\begin{equation}
p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) = p(z) [z \in A] \alpha(z) \mbox{ for all } z \in Z
\end{equation}
The following choice of $\alpha$ suffices:
\begin{equation}
\alpha(z) = \min\left\{ 1, \frac{p(h(z))}{p(z)} \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \right\}
\end{equation}

@@

>>>>

% graphics
\usepackage{float}
\usepackage{graphicx}

@@

\begin{abstract}
Involutive MCMC is a unifying framework for MCMC that encompasses many standard
and recent MCMC algorithms from the literature, from reversible jump MCMC to
Hamiltonian Monte Carlo to kernels based on deep neural networks. The key idea
in involutive MCMC is that a combination of (i) an auxiliary probability
distribution and (ii) an involution on the extended state space that includes
the model latent variables and the auxiliary variables results in MCMC kernels
that satisfy detailed balance with respect to the target distribution. The
involutive MCMC framework is appealing for its simplicity and generality, and
promises to be a useful conceptual and mathematical tool for exploring the
design space of MCMC kernels. However, like other Monte Carlo samplers,
implementing involutive samplers is time consuming and error prone. This paper
describes a technique for automatically generating the implementation of
involutive MCMC samplers from two probabilistic programs that define the target
distribution and the auxiliary probability distribution respectively, and a
differentiable program that defines the involution. The technique also detects
common conceptual and programming errors that arise when designing and
specifying involutive MCMC algorithms. This paper describes the support for
involutive MCMC in the Gen probabilistic system that uses this technique, and
shows involutive samplers specified using Gen's high-level probabilistic and
differentiable programming languages.
\end{abstract}

@@

%This is the best paper~\citep{cusumano2019gen}.

@@

\input{headline-figure}

@@

% TODO can only show this for a special case -- discrete involution + continuous?, or some more general case?

% TODO show an performance table of moves per second with and without the sparsity reduction (?)

@@

% TODO show an example of the type of bug it can detect (an edge case?)

@@

\subsection{Distributions on traces}
Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.

% TODO rename the joint target density to \pi below
We want to show stationarity with respect to $\pi$, using stationarity of the involution with respect to the joint density (below).
\begin{equation}
\int_X p(x) \ell_x(A) \mu(dx) = p(A) \mbox{ for all } A \in \Sigma
\end{equation}
where $\ell_x(A)$ is the probability of transitioning into set $A$ from $x$, and is given by:
\begin{equation}
\ell_x(A) := \int_U k_{(x, u)}(A \times U) \mu(du)
\end{equation}
where $k_{(x, u)}$ is the measure on $X \times U$ that is induced by applying the involution $h$ to the state $(x, u)$.
% TODO checkme

\subsection{Stationarity of the involution with respect to joint density}
Let $(Z, \Sigma, \mu)$ be a measure space of pairs of traces of the model and proposal, where $\mu$ is $\sigma$-finite.
Consider a $\mu$-measurable function $h : Z \to Z$ that is an involution ($h(h(z)) = z$) and let $\mu \circ h^{-1}$ denote the pushforward of $\mu$ under $h$, such that $\mu \circ h^{-1}$ is $\sigma$-finite and such that $\mu \circ h^{-1}$ is absolutely continuous with respect to $\mu$.
% TODO can we simplify the conditions, using the fact that h is an involution?
Then, the Radon-Nikodym derivative exists, and is denoted $(d (\mu \circ h^{-1}) / d \mu)(z)$.

The goal is to prove stationarity with respect to the measure defined by density $p(z)$ with respect to $\mu$.
The move starts with state $z \in Z$, applies the involution $h$, and then accepts (returns $h(z)$) with probability $\alpha(z)$ and otherwise returns $z$.
Let $k_{z}$ denote the measure on new states induced by the involution $h$ and the previous state $z$:
\begin{equation}
k_{z}(A) = [h(z) \in A] \alpha(z) + [z \in A] (1 - \alpha(z))
\end{equation}
The stationarity condition is:
\begin{equation}
\int_Z p(z) k_{z}(A) \mu(dz) = p(A) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
Substituting in the definition of $k_z(A)$ and simplifying gives:
\begin{equation} \label{eq:stationarity-requirement}
\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z p(z) [z \in A] \alpha(z) \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
Note that for a $\mu$-measurable function $g$ such that $g \circ h$ is integrable with respect to $\mu$:
\begin{equation} \label{eq:pushforward}
\int_Z g(h(z)) \mu(dz) = \int_Z g(z) (\mu \circ h^{-1})(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
\end{equation}
For $g(z) := p(h(z)) [z \in A] \alpha(h(z))$ we have $g(h(z)) = p(z) [h(z) \in A] \alpha(z)$, which is the integrand of the left-hand-side of Equation~(\ref{eq:stationarity-requirement}).
Applying Equation~(\ref{eq:pushforward}) to the left-hand side of Equation~(\ref{eq:stationarity-requirement}) gives:
\begin{equation}
\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
\end{equation}
Equating this with the right-hand side of Equation~(\ref{eq:stationarity-requirement}) gives the following sufficient condition for stationarity:
\begin{equation}
\int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
It therefore suffices to find $\alpha$ such that:
\begin{equation}
p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) = p(z) [z \in A] \alpha(z) \mbox{ for all } z \in Z
\end{equation}
The following choice of $\alpha$ suffices:
\begin{equation}
\alpha(z) = \min\left\{ 1, \frac{p(h(z))}{p(z)} \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \right\}
\end{equation}

@@

\paragraph{Characterizing the pushforward $\mu \circ h^{-1}$ for involution $h$}
Can we guarantee that it is $\sigma$-finite or absolutely continuous with respect to $\mu$?
Are there simpler sufficient conditions we can give?
$(\mu \circ h^{-1})$ is $\sigma$-finite if $Z$ is the countable union of $(mu \circ h^{-1})$-measurable sets with finite measure.

$\sigma$-finite means
